Solving the Poisson-equation in one dimension: \\ \normalsize{TDMA vs LU} \\ \hrulefill\small{ FYS3150: Computational Physics }\hrulefill

### 2016-09-16 Project1/article/article.tex
A final read-through?

\section{Theory}
\subsection{The Poisson Equation}
The Poisson equation is a classical equation from electromagnetism. The electrostatic potential $\Phi$ is generated by a localized charge distribution $\rho(\vb{r})$. In three dimensions the equation reads
\begin{equation}
\laplacian\Phi=-4\pi \rho(\vb{r})
\end{equation}
where $\laplacian$ is the Laplace operator. In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study I am assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem. only dependent on radius $r$.
\begin{equation}
\laplacian=\frac{1}{r^2}\frac{d}{dr}\left(r^2\frac{d\Phi}{dr} \right)
\end{equation}
By substituting $\Phi(r)=\phi(r)/r$ the Poisson equation is reduced to 
\begin{equation}
\frac{d^2\phi}{dr^2}=-4\pi r\rho(r)
\end{equation}
and by letting $\phi \rightarrow u$ and $r \rightarrow x$ one is left with the very simple equation
\begin{equation}
-u''(x)=f(x) \label{eq:2nd}
\end{equation}

@@

\section{Algorithms}
Two main methods are implemented and compared. The first method is
gaussian elimination of the tridiagonal matrix $A$, also known as the
\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
elimination that can ve used to solve tridiagonal systems of
equations. The method is improved upon in order to take into account
the fact that the matrix we are dealing with has the same numbers
along the diagonals. The second method is the LU-decomposistion
method. Code snippets are included in this section, for the full programs follow the link to github at the first page of this article.

@@

The Gaussian elimination is not fully complete until a ``backward
substitution'' has been performed as well. This ensures that all
eliments in the row reduced matrix are pivot elements.

@@

By looking at the results one is led to believe that there are more reasons to prefer the tridiagonal matrix algorithm for this kind of problem. When using the LU decomposition method the short term memory is quickly filled with nothing but the matrix $A$. Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes$=80$GB of memory! A brand new computer ships with around $8$ GB of memory today, which is a tenth of the space required. Obviously, this inhibits our ability to perform the calculation.

>>>>

\section{Theory}
\subsection{The Poisson Equation}
The Poisson equation is a classical equation from electromagnetism. The electrostatic potential $\Phi$ is generated by a localized charge distribution $\rho(\vb{r})$. In three dimensions the equation reads
\begin{equation}
\laplacian\Phi=-4\pi \rho(\vb{r})
\end{equation}
where $\laplacian$ is the Laplace operator. In three dimensions the Laplace operator can be expressed using spherical coordinates, but in this study we are assuming that $\Phi$ and $\rho$ are spherically symmetric, thus reducing the equation to a one-dimensional problem. only dependent on radius $r$.
\begin{equation}
\laplacian=\frac{1}{r^2}\frac{d}{dr}\left(r^2\frac{d\Phi}{dr} \right)
\end{equation}
By substituting $\Phi(r)=\phi(r)/r$ the Poisson equation is reduced to 
\begin{equation}
\frac{d^2\phi}{dr^2}=-4\pi r\rho(r)
\end{equation}
and by letting $\phi \rightarrow u$ and $r \rightarrow x$ one is left with the very simple equation
\begin{equation}
-u''(x)=f(x) \label{eq:2nd}
\end{equation}

@@

\section{Algorithms}
Two main methods are implemented and compared. The first method is
gaussian elimination of the tridiagonal matrix $A$, also known as the
\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
elimination that can be used to solve tridiagonal systems of
equations. The method is improved upon in order to take into account
the fact that the matrix we are dealing with has the same numbers
along the diagonals. The second method is the LU-decomposistion
method. Code snippets are included in this section, for the full programs, follow the link to github at the first page of this article.

@@

The Gaussian elimination is not fully complete until a ``backward
substitution'' has been performed as well. This ensures that all
elements in the row reduced matrix are pivot elements.

@@

By looking at the results one is led to believe that there are more reasons to prefer the tridiagonal matrix algorithm for this kind of problem. When using the LU decomposition method the short term memory is quickly filled with nothing but the matrix $A$. Choosing $n=10^5$ is equivalent to $8\times10^{10}$ bytes $=80$GB of memory! A brand new computer ships with around $8$ GB of memory today, which is a tenth of the space required. Obviously, this inhibits our ability to perform the calculation.
